Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-18 paper/paper.tex
Changes

%% WARNING: removing the point about axiom K leads to leaving the use of
%% decidable validity predicates unexplained -- do not remove!
We define containers in terms of shapes, \emph{base} positions
(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
base position is valid in a shape. Since validity predicates are in \coq{Prop},
they will be erased during extraction, and since they are decidable, they
prevent the use of axiom K to deal with heterogeneous equalities of positions:
\begin{coqcode}
Record App F (X : Type) := {shape : Shape F; cont : {p : APos | valid s p = true} -> X}.
\end{coqcode}

@@

Recall that an algebra is a set $A$ together with a morphism that defines the
operations of the algebra $F\;A \to A$.  Given a type \coq{A} and a container
\coq{F}, an `\coq{App F}'-algebra is a pair given by the carrier \coq{A}, and the
\emph{structure map} of type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}

\noindent
For simplicity, we will abuse terminology, and call these
\coq{F}-algebras, instead of `\coq{App F}'-algebras.

The initial \coq{F}-algebra can be defined as follows, thanks to the use of
containers:
\begin{coqcode}
Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}

\noindent
where \coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
initial \coq{F}-algebra for the container that is isomorpic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra being
defined by the empty list \coq{Empty : unit -> LFix F} and the cons operation
\coq{Cons : A * LFix F -> LFix F}.

We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR} (see
Section~\ref{sec:types-morphisms}). We define smart constructors for the
isomorphism of least fixed points as respectful morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}

Catamorphisms are constructed, as expected, so that they structurally
deconstruct the datatype, call themselves recursively, and then compose the
result using an \coq{F}-algebra.
\begin{coqcode}
Definition cata_f (alg : Alg F A) : LFix -> A
  := fix f (x : LFix)
     := match x with
        | LFix_in ax =>
             alg {shape := shape ax; cont := fun e => f (cont ax e)}
        end.
\end{coqcode}

It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}

>>>>

%% WARNING: removing the point about axiom K leads to leaving the use of
%% decidable validity predicates unexplained -- do not remove!
We define containers in terms of shapes, \emph{base} positions
(\coq{APos : Set}), and \emph{decidable validity predicates} that specify when a
base position is valid in a shape. Since validity predicates are in \coq{Prop},
they will be erased during extraction, and since they are decidable, they
prevent the use of axiom K to deal with heterogeneous equalities of positions:
\begin{coqcode}
Record App F X := {shape : Shape F; cont : {p : APos F | valid s p = true} -> X}.
\end{coqcode}

@@

Recall that an algebra is a set $A$ together with a morphism that defines the
operations of the algebra $F\;A \to A$.  Given a type \coq{A} and a container
\coq{F}, an `\coq{App F}'-algebra (or `\coq{F}-algebra' for short) is a pair
given by the carrier \coq{A}, and the \emph{structure map} of type:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The initial \coq{F}-algebra can be defined as the inductive type with a finite
number of uses of \coq{App F}
\begin{coqcode}
Inductive LFix F : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
\coq{LFix_in} is the initial algebra $\InIso$, while \coq{LFix_out} is its
inverse $\InOp$ (see Section~\ref{sec:category-theory}).  As an example, the
initial \coq{F}-algebra for the container that is isomorpic to the functor
\coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra being
defined by the empty list \coq{Empty : unit -> LFix F} and the cons operation
\coq{Cons : A * LFix F -> LFix F}.

The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
setoid equality. We define smart constructors for the isomorphism of least fixed
points as proper morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}

Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}

It is easy to show that this function is a proper morphism of \coq{F}-algebras.
In fact, it is possible to define it as a map of the following type:
\begin{coqcode}
  cata : forall `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
